Visualizing Independence Using Extended Association Plots

### 2005-08-04 trunk/vcd/paper/JCGS.Rnw
sections 4 and 5 finished, only intro and abstract missing
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@435 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

  set.seed(myseed)
  ucb_cotab <- cotab_coindep(ucb, condvars = "Dept", type = "assoc", n = 5000)

@@

\noindent \textbf{Key Words:} Association plots; Conditional inference; 
Contingency tables; HCL colors; HSV colors; Mosaic plots; Permutation tests.
\end{abstract}

@@

To fix notations, we consider a 2-way contingency table with cell
frequencies $\{n_{ij}\}$ for $i = 1, \dots, I$ and $j = 1, \dots, J$
and row and column sums $n_{i+} = \sum_i n_{ij}$ and $n_{+j} = \sum_j n_{ij}$
respectively.
For convenience the number of observations is denoted $n = n_{++}$.
Given an underlying distribution with theoretical cell probabilities
$\pi_{ij}$, the null hypothesis of independence of the two categorical
variables can be formulated as
\begin{equation} \label{eq:H0}
H_0: \; \pi_{ij} \quad = \quad \pi_{i+} \pi_{+j}.
\end{equation}

The expected cell frequencies in this model are
$\hat n_{ij} = n_{i+}n_{+j}/n$. The probably best known and most used
measure of discrepancy between observed and expected values are the Pearson residuals
\begin{equation} \label{eq:Pearson}
r_{ij} \quad = \quad \frac{n_{ij} - \hat n_{ij}}{\sqrt{\hat n_{ij}}}.
\end{equation}
The most convenient way to aggregate the $I \times J$ residuals
to one test statistic is their squared sum
\begin{equation} \label{eq:ChiSquare}
X^2 \quad = \quad \sum_{i, j}r_{ij}^2,
\end{equation}
because this is known to have an unconditional limiting $\chi^2$ distribution with $(I-1)(J-1)$
degrees of freedom under the null hypothesis.
This is the well-known $\chi^2$ test which is typically introduced at the very beginning
of the chapter about independence in 2-way tables in statistics textbooks
\citep[see e.g.,][]{vcd:Agresti:2002}.

@@

But the sum of squares is not the only plausible way of capturing deviations
from $0$ in the Pearson residuals.
There are many other conceivable functionals $\lambda(\cdot)$ which
lead to reasonable test statistics $\lambda(\{r_{ij}\})$: one which is particularly
suitable for identifying the cells which cause the dependence (if any) is
the maximum of the absolute values
\begin{equation} \label{eq:MaxAbs}
M \quad = \quad \max_{i, j} |r_{ij}|.
\end{equation}
Given a critical value $c_\alpha$ for this test statistic, all residuals whose
absolute value exceeds $c_\alpha$ violate the hypothesis of indendence at significance level
$\alpha$ \cite[][Ch.~7]{vcd:Mazanec+Strasser:2000}. Thus, the interesting cells causing
the dependence can easily be identified.

Furthermore, an important reason for using the unconditional limiting distribution
for the $X^2$ statistic from (\ref{eq:ChiSquare}) was the closed form result
for the distribution. Recently, with the improving perfomance of computers,
conditional inference (or permutation tests)---carried out either by simulation
or by computation of the (asymptotic) permutation distribution---have been receiving
increasing attention \citep{vcd:Pesarin:2001,vcd:Strasser+Weber:1999}. 
For testing the independence hypothesis from (\ref{eq:H0}), using a permutation test is particularly
intuitive due to the permutation invariance (given row and column sums) of this problem.
Consequently, all results in this paper are based on conditional inference performed
by simulating from the permutation distribution of test statistics of type
$\lambda(\{r_{ij}\})$.

Other measures of discrepancy (e.g., deviance residuals) could, of course, also be used
instead of the the Pearson residuals $\{r_{ij}\}$. But as the ideas discussed here
extend straightforwardly to that situation, we do not discuss it in detail.

@@

Two well-established visualization techniques for independence in 2-way tables are
mosaic plots and association plots. Both are suitable to bring out
departures of an observed table $\{n_{ij}\}$ from the expected table $\{\hat n_{ij}\}$
in a graphical way. The latter focuses on the visualization of the Pearson residuals
$r_{ij}$ (under independence) while the latter primarily displays the observed frequencies
$n_{ij}$.

@@

Colors are commonly used to enhance mosaic and association plots. To integrate
a visualization of the Pearson residuals $\{r_{ij}\}$ into the mosaic display---which
in its `raw' version only visualizes the observed frequencies $\{n_{ij}\}$---\cite{vcd:Friendly:1994}
suggested a residual-based shading for the mosaic tiles that can also be applied
to the rectangles in association plots \citep{vcd:Meyer+Zeileis+Hornik:2003}. 
In this section, we first briefly review the \cite{vcd:Friendly:1994} shading, 
before we suggest different colors and a combination of visualization and significance
testing to extend these residual-based shadings.

@@

The idea is to use a color coding for the mosaic tiles that visualizes the
sign and absolute size of each residual $r_{ij}$: Cells corresponding to small
residuals ($|r_{ij}| < 2$) are shaded white. Cells with medium sized residuals
($2 \le |r_{ij}| < 4$) are shaded in light blue and light red for positive negative
residuals respectively. Cells with large residuals ($|r_{ij}| \ge 4$) are shaded
with a fully saturated blue and red respectively. Mosaic plots enhanced by this 
shading can thus also bring out departures from independence (or other log-liner 
models in multi-way tables) graphically and visualize patterns of dependence.
The heuristic for choosing the cut offs $2$ and $4$ is that the Pearson residuals
are approximately standard normal which implies that the highlighted cells are those with
residuals \emph{individually} significant at approximately the $\alpha = 0.05$ and $\alpha = 0.0001$
level. But the main purpose of the shading is not to visualize significance but the \emph{pattern} of
deviation from independence \citep[p.~109]{vcd:Friendly:2000}.

@@

The HSV space originally looks like a regular cone with black at its
peak (zero value) and full color wheels for different saturations at the other end,
centered around a white (full value). Type and amount of color are controlled by hue
and saturation respectively. Typically, polar coordinates that all range in the 
unit interval are used in this space, such that it looks like a regular cylinder.
For generating colors in the Friendly shading, the following strategy is used:
The hue $h$ codes the sign of the residuals---$h = 0$ (red hue) is used for negative residuals,
$h = 2/3$ (blue hue) for positive residuals. The absolute size of the residuals
is then coded by the saturation $s$ which is set to $0$, $0.5$ and $1$ respectively
for small/medium/large residuals respectively. The value is always fixed at $v = 1$.
This is also depicted in the upper panel of Figure~\ref{fig:shadingHSVHCL} which shows
to slides from the HSV cylinder side to side. The plot shows the saturation/value plane
for the given hues $h = 0$ and $h = 2/3$. The significant color palette shows the
colors used for the Friendly shading when the residuals are increasing from left to right.
The set of non-significant colors will be explained in the next section.

Although this HSV-based shading is already very useful for enhancing mosaic and
association plots and although HSV is very commonly available implementation of
color spaces, HSV colors in general and the Friendly shading in particular have
a number of disadvantages. Most importantly, HSV colors are not perceptually
uniform because the three HSV dimensions map only poorly to the three
perceptual dimensions of the human visual system  \citep{vcd:Brewer:1999,vcd:Ihaka:2003}.
Consequently, the the HSV dimensions are confounded, e.g., saturation
is not uniform across different hues. A fully saturated blue $(2/3, 1, 1)$ is
much darker as a fully saturated red $(0, 1, 1)$ or green $(1/3, 1, 1)$.
This makes it more difficult for the human eye to judge the size of shaded areas
and can therefore lead to color-caused optical illusions when used in statistical graphs
\citep{vcd:Cleveland+McGill:1983}. Furthermore, flashy fully saturated HSV colors are
good for drawing attention to a plot, but hard to look at for a longer time \citep{vcd:Ihaka:2003}
which makes HSV-shaded graphics harder to interpret. Finally, white is employed as
the neutral color for small residuals in the Friendly shading, but typically
grey is found to convey neutrality or uninterestingness much better than white does
\citep{vcd:Brewer:1999}.

Alternative ways to choose colors have been available for a long time, but have
been only slowly adopted for implementations of colors in computer packages in 
general and for shading in statistical graphs in particular. The idea of using
perceptually based colors that are `in harmony' go back to \cite{vcd:Munsell:1905} 
who introduced a color notation for balanced colors. Based on similar principles, Cynthia
Brewer and co-workers suggested different types of palettes (qualitative/sequential/diverging)
and provided the online tool \pkg{ColorBrewer.org} \citep{vcd:Harrower+Brewer:2003} for
selecting an appropriate palette for a specific problem. Furthermore, the
Commission Internationale de l'\'Eclairage (CIE) introduced the two perceptually
based color spaces CIELAB and CIELUV where the latter is typically preferred for 
emissive color technologies such as computer displays. \cite{vcd:Ihaka:2003} discusses
how CIELUV colors can be used for choosing colors for statistical graphics such as barplots.
By taking polar coordinates in CIELUV space, it is called HCL (Hue--Chroma--Luminance) space
and qualitative palettes can easily be chosen by using a range of hues for fixed values
of chroma and luminance. Such colors are always balanced towards the same grey 
and thus do not have the problem of varying saturations as for the HSV colors.

Here, we discuss how similar ideas can be used for deriving a diverging HCL palettes that 
provide a suitable translation of the ideas from the Friendly shading to perceptually
uniform HCL colors. 
The HCL space looks like a double cone with black (zero luminance) at one end and white
at the other (full luminance). In its middle, there is a full color wheel for different
values of chroma (that controls the colorfulness). Unfortunately, the HCL space is not completely
regular as the HSV space is and consequently its dimensions cannot be standardized to 
unit intervals: the hue ranges in $[0, 360]$ degrees and chroma and luminance both in $[0, 100]$
percent. But not all combinations $(h, c, l)$ yield valid HCL colors and the admissable combinations
of $c$ and $l$ vary across different hues $h$. But for the problem of constructing
a diverging palette, this problem can easily be overcome as we just need two different
hues (a `negative' and a `positive' hue) and hence choose two hues that correspond to 
similar shapes in the chroma/luminance plane. The lower panel of Figure~\ref{fig:shadingHSVHCL}
shows two such planes side to side for the hues $h = 0$ and $h = 260$. To obtain a 
sequence of colors with the same properties as the Friendly shading, the palette starts
at a fully saturated red $(0, 100, 50)$, goes via a neutral color, ends at a fully saturated
blue $(260, 100, 50)$, and uses linear interpolation in between. Instead of using white
$(0, 0, 100)$ as the neutral color, a light grey $(0, 0, 90)$ is employed as motivated above.

This diverging palette (see the `significant' colors in Figure~\ref{fig:shadingHSVHCL},
the `non-significant' colors are again discussed in the following section)
uses both chroma, i.e., the colorfulness, and luminance, i.e., 
the amount of grey, to code the absolute size of the quantity visualized, i.e., the
Pearson residuals when applied to the independence problem. By changing the neutral
color or by changing the maximum chroma respectively, this can be changed to using only
chroma or luminance for this purpose, but using both seems to be a very effective way
of visualization.

The advantages of using this HCL shading scheme instead of the HSV scheme is that
it is based on a perceptually based color space that provides uniform saturations
for different hues and is device inependent.

@@

The shading scheme of \cite{vcd:Friendly:1994} was suggested to visualize
the pattern of dependence in contingency tables, as discussed above, but
the presence (or absence) of colors in a plot also always conveys an
impression of interestingness (or uninterstingness respectively). That is,
statisticians might be tempted to interpret the absence of color in a plot
as a clue that there is no significant departure from independence. Or vice
versa, colored cells would convey the impression that there is significant
dependence. Currently, both is not true as can be seen in the left panel
of Figure~\ref{fig:shadings} which shows the mosaic display for the arthritis
data with Friendly shading. Although there is significant dependence, no 
residual exceeds an absolute value of $2$ and hence no cell is colored.
Of course, it can be argued that the shading was not designed for this purpose
and that different cut offs than $2$ and $4$ should be used here. But then
again, it would be nice if these cut offs could be chosen automatically in a
data-driven fashion which is what we do below.

The Friendly shading can be interpreted to be a visualization of the maximum
statistic $M$ from (\ref{eq:MaxAbs}) which always employs the critical values $c_\alpha$
of $2$ and $4$ respectively. But the problem is that it is not clear to which
significance levels $\alpha$ these critical values correspond because the distribution
of $M$ depends on the underlying contingency table. The natural solution to this
problem is to compute the critical values from the distribution of $M$ in a data-driven
way (i.e., for the table visualized) and use these instead of the hard-coded values
$2$ and $4$. In the right panel of Figure~\ref{fig:shadings} this is done for
the arthritis data by employing the critical values
$\Sexpr{round(as.vector(art_max$qdist(0.9)), digits = 2)}$ at level $\alpha = 0.1$ and
$\Sexpr{round(as.vector(art_max$qdist(0.99)), digits = 2)}$ at level $\alpha = 0.01$
(and using the diverging HCL palette) derived from the permutation distribution of $M$ for 
the arthritis data as described in Section~\ref{sec:tests}. By using these cut offs, the presence of color
in the plot is equivalent to significance (of the maximum statistic $M$) at 
level $\alpha = 0.1$ and $\alpha = 0.01$ respectively and exactly the cells which `cause'
the dependence are highlighted. For the arthritis data, these are in particular the 
cells in the last column that signal that there are significantly more
treated patients and fewer placebo patients with marked improvement than would be expected
under independence.

The significance $\alpha = 0.1$ and $0.01$ are chosen because this leads to displays where
fully colored cells are clearly significant (i.e., with a $p$ value below $0.01$),
cells without color are clearly non-significant (i.e., with a $p$ value above $0.1$), and
cells in between can be considered to be weakly significant (i.e., $0.1 \le p < 0.01$).
Of course, users could choose any other set of significance levels they feel comfortable
with, e.g., only a single cutoff at $\alpha = 0.05$ or three cut offs at $0.1$, $0.05$ and $0.01$
etc. Another option could be to use a continuous shading where the $p$ value corresponding
to a cell controls the interpolation between the neutral and the full color. However, our
experience is that this typically results in too much color in a plot which in turn tends
to conceal the important cells and over-emphasizes the unimportant ones. Hence, we stick
to a shading with fewer colors that are easier to interpret.

This extended shading is already very flexible and combines visualization and inference.
However, it can only be applied when employing the maximum statistic because it is the only
aggregation functional $\lambda(\cdot)$ where a single large residual $|r_{ij}|$ exceeding
its critical value (independent from the values of the remaining residuals) is equivalent
to the a significant value of the whole test statistic $\lambda(\{r_{ij}\})$.\footnote{This
can easily be seen from the following consideration: To obtain a level $\alpha$ test, where 
any residual $|r_{ij}|$ exceeding the critical value $c_\alpha$ gives a significant test
result, the equation $\alpha = P(\mbox{any } |r_{ij}| > c_\alpha)$ must hold. This is equivalent
to $1 - \alpha = P(\mbox{all } |r_{ij}| \le c_alpha) = P(\max_{i,j} |r_{ij}| \le c_alpha)$.}
Typically, applying the maximum statistic is feasible and also appropriate for exploratory 
analysis, but it would be desirable to also have a residual-based shading that can incorporate
visualization of significance when the sum-of-squares statistic $X^2$ (or any other functional
$\lambda(\cdot)$) is used. For the reasons discussed above, it is not possible to achieve this by shading
individual cells differently but can only be realized by using different colors for the
whole table. A outlined before, colorfulness is often intuitively matched with interestingness,
therefore a rather natural idea is to use the fully colored palette only when the corresponding
test is significant and to use a less colorful palette if not. For the HCL scheme, the
amount of color can easily be controlled by varying the maximum chroma value used. For the full
colors, the maximum chroma was set to $100$ as shown in Figure~\ref{fig:shadingHSVHCL} and 
is now reduced to $20$ for the non-significant palette. This palette still codes the absolute
size of the residuals by luminance (i.e., the amount of grey), uses the same neutral grey
for small `un-interesting' residuals, codes positive and negative residuals by different hues,
but gives less emphasis to the pattern by making the plot less colorful. A similar effect
can be obtained in the HSV space if the value is reduced from $1$ to $0.5$ for non-significant 
palettes (see Figure~\ref{fig:shadingHSVHCL}). But as for the significant colors,
the dimensions used for creating this palette are confounded and hence the HCL scheme is clearly
preferable.

\setkeys{Gin}{width=\textwidth}
\begin{figure}[htbp]
\begin{center}
<<shadings,fig=TRUE,echo=FALSE,height=6,width=11,results=hide>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
mosaic(art, newpage = FALSE, gp = shading_Friendly, margins = c(2.5, 4, 2.5, 3), gp_args = list(eps = NULL, lty = 1))
popViewport()
pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
mosaic(art, newpage = FALSE, gp = art_gp, margins = c(2.5, 3, 2.5, 4))
popViewport(2)
@
\caption{Mosaic plot for the Arthritis data with Friendly shading (left) and extended shading (right).}
\label{fig:shadings}
\end{center}
\end{figure}

\section{Extensions to Conditional Independence}

UCB admissions data \citep{vcd:Bickel+Hammel+O'Connell:1975}
double max: \Sexpr{stat(ucb_doublemax)}, p-value = \Sexpr{pval(ucb_doublemax)}
max Chisq: \Sexpr{stat(ucb_maxChisq)}, p-value = \Sexpr{pval(ucb_maxChisq)}
sum Chisq: \Sexpr{stat(ucb_sumChisq)}, p-value = \Sexpr{pval(ucb_sumChisq)}
simulated from $5000$ repetitions

\cite{vcd:Friendly:1999} discusses grouping similar to coplots (conditioning
plots) \citep{vcd:Cleveland:1993} that lead to trellis graphics
\citep{vcd:Becker+Cleveland+Shyu:1996}.

>>>>

  set.seed(myseed)
  ucb_cotab <- cotab_coindep(ucb, condvars = "Dept", type = "assoc", n = 5000, margins = c(3, 1, 1, 3))

@@

\noindent \textbf{Key Words:} Association plots; Conditional inference; 
Contingency tables; HCL colors; HSV colors; Mosaic plots.
\end{abstract}

@@

To fix notations, we consider a 2-way contingency table with cell
frequencies $\{n_{ij}\}$ for $i = 1, \dots, I$ and $j = 1, \dots, J$
and row and column sums $n_{i+} = \sum_i n_{ij}$ and $n_{+j} = \sum_j n_{ij}$
respectively.
Given an underlying distribution with theoretical cell probabilities
$\pi_{ij}$, the null hypothesis of independence of the two categorical
variables can be formulated as
\begin{equation} \label{eq:H0}
H_0: \; \pi_{ij} \quad = \quad \pi_{i+} \pi_{+j}.
\end{equation}

The expected cell frequencies in this model are
$\hat n_{ij} = n_{i+}n_{+j}/n_{++}$. The probably best known and most used
measure of discrepancy between observed and expected values are the Pearson residuals
\begin{equation} \label{eq:Pearson}
r_{ij} \quad = \quad \frac{n_{ij} - \hat n_{ij}}{\sqrt{\hat n_{ij}}}.
\end{equation}
The most convenient way to aggregate the $I \times J$ residuals
to one test statistic is their squared sum
\begin{equation} \label{eq:ChiSquare}
X^2 \quad = \quad \sum_{i, j}r_{ij}^2,
\end{equation}
because this is known to have an unconditional limiting $\chi^2$ distribution with $(I-1)(J-1)$
degrees of freedom under the null hypothesis.
This is the well-known $\chi^2$ test which is typically introduced at the very beginning
of the chapter about independence in 2-way tables in statistics textbooks
\citep[see e.g.,][]{vcd:Agresti:2002}.

@@

But the sum of squares is not the only plausible way of capturing deviations
from $0$ in the residuals.
There are many other conceivable functionals $\lambda(\cdot)$ which
lead to reasonable test statistics $\lambda(\{r_{ij}\})$: one which is particularly
suitable for identifying the cells which cause the dependence (if any) is
the maximum of the absolute values
\begin{equation} \label{eq:MaxAbs}
M \quad = \quad \max_{i, j} |r_{ij}|.
\end{equation}
Given a critical value $c_\alpha$ for this test statistic, all residuals whose
absolute value exceeds $c_\alpha$ violate the hypothesis of indendence at significance level
$\alpha$ \cite[][ch.~7]{vcd:Mazanec+Strasser:2000}. Thus, the interesting cells causing
the dependence can easily be identified.

Furthermore, an important reason for using the unconditional limiting distribution
for the $X^2$ statistic from (\ref{eq:ChiSquare}) was the closed form result
for the distribution. Recently, with the improving perfomance of computers,
conditional inference (or permutation tests)---carried out either by simulation
or by computation of the (asymptotic) permutation distribution---have been receiving
increasing attention \citep{vcd:Pesarin:2001,vcd:Strasser+Weber:1999}. 
For testing the independence hypothesis from (\ref{eq:H0}), using a permutation test is particularly
intuitive due to the permutation invariance (given row and column sums) of this problem.
Consequently, all results in this paper are based on conditional inference performed
by simulating the permutation distribution of test statistics of type
$\lambda(\{r_{ij}\})$.

Other measures of discrepancy (e.g., deviance residuals) could, of course, also be used
instead of the the Pearson residuals $\{r_{ij}\}$. But as the ideas discussed here
extend straightforwardly to that situation, we do not go into detail about this.

@@

Two well-established visualization techniques for independence in 2-way tables are
mosaic plots and association plots. Both are suitable to bring out
departures of an observed table $\{n_{ij}\}$ from the expected table $\{\hat n_{ij}\}$
in a graphical way. The latter focuses on the visualization of the Pearson residuals
$r_{ij}$ (under independence) while the former primarily displays the observed frequencies
$n_{ij}$.

@@

Colors are commonly used to enhance mosaic and association plots. To integrate
a visualization of the residuals $\{r_{ij}\}$ into the mosaic display---which
in its `raw' version only visualizes the observed frequencies $\{n_{ij}\}$---\cite{vcd:Friendly:1994}
suggested a residual-based shading for the mosaic tiles that can also be applied
to the rectangles in association plots \citep{vcd:Meyer+Zeileis+Hornik:2003}. 
In this section, we first briefly review the \cite{vcd:Friendly:1994} shading, 
before we suggest different colors and a combination of visualization and significance
testing to extend these residual-based shadings.

@@

The idea is to use a color coding for the mosaic tiles that visualizes the
sign and absolute size of each residual $r_{ij}$: Cells corresponding to small
residuals ($|r_{ij}| < 2$) are shaded white. Cells with medium sized residuals
($2 \le |r_{ij}| < 4$) are shaded light blue and light red for positive and negative
residuals respectively. Cells with large residuals ($|r_{ij}| \ge 4$) are shaded
with a fully saturated blue and red respectively. Mosaic plots enhanced by this 
shading can thus also bring out departures from independence (or other log-linear 
models in multi-way tables) graphically and visualize patterns of dependence.
The heuristic for choosing the cut offs $2$ and $4$ is that the Pearson residuals
are approximately standard normal which implies that the highlighted cells are those with
residuals \emph{individually} significant at approximately the $\alpha = 0.05$ and $\alpha = 0.0001$
level. But the main purpose of the shading is not to visualize significance but the \emph{pattern} of
deviation from independence \citep[p.~109]{vcd:Friendly:2000}.

@@

The HSV space originally looks like a regular cone with black at its
peak (zero value) and full color wheels for different saturations at the other end,
centered around a white (full value). Type and amount of color are controlled by hue
and saturation respectively. Typically, polar coordinates that all range in the 
unit interval are used in this space, such that it looks like a regular cylinder.
For generating colors in the Friendly shading, the following strategy is used:
The hue $h$ codes the sign of the residuals---$h = 0$ (red hue) is used for negative residuals,
$h = 2/3$ (blue hue) for positive residuals. The absolute size of the residuals
is then coded by the saturation $s$ which is set to $0$, $0.5$ and $1$ respectively
for small/medium/large residuals. The value is always fixed at $v = 1$.
This is also depicted in the upper panel of Figure~\ref{fig:shadingHSVHCL} which shows
two slides from the HSV cylinder side by side. The plot shows the saturation/value plane
for the given hues $h = 0$ and $h = 2/3$. The significant color palette shows the
colors used for the Friendly shading when the residuals are increasing from left to right.
The set of non-significant colors will be explained in the next section.

Although this HSV-based shading is already very useful for enhancing mosaic and
association plots and although HSV is very commonly available implementation of
color spaces, HSV colors in general and the Friendly shading in particular have
a number of disadvantages. Most importantly, HSV colors are not perceptually
uniform because the three HSV dimensions map only poorly to the three
perceptual dimensions of the human visual system  \citep{vcd:Brewer:1999,vcd:Ihaka:2003}.
Consequently, the HSV dimensions are confounded, e.g., saturation
is not uniform across different hues. A fully saturated blue $(2/3, 1, 1)$ is
perceived to be much darker than a fully saturated red $(0, 1, 1)$ or green $(1/3, 1, 1)$.
This makes it more difficult for the human eye to judge the size of shaded areas
and can therefore lead to color-caused optical illusions when used in statistical graphs
\citep{vcd:Cleveland+McGill:1983}. Furthermore, flashy fully saturated HSV colors are
good for drawing attention to a plot, but hard to look at for a longer time \citep{vcd:Ihaka:2003}
which makes HSV-shaded graphics harder to interpret. Finally, white is employed as
the neutral color for small residuals in the Friendly shading, but typically
grey is found to convey neutrality or un-interestingness much better than white
\citep{vcd:Brewer:1999}.

Alternative ways to choose colors have been available for a long time, but have
been only slowly adopted for implementations of colors in computer packages in 
general and for shading in statistical graphs in particular. The idea of using
perceptually based colors that are `in harmony' go back to \cite{vcd:Munsell:1905} 
who introduced a color notation for balanced colors. Based on similar principles, Cynthia
Brewer and co-workers suggested different types of palettes (qualitative/sequential/diverging)
and provided the online tool \pkg{ColorBrewer.org} \citep{vcd:Harrower+Brewer:2003} for
selecting an appropriate palette for a specific problem. Furthermore, the
Commission Internationale de l'\'Eclairage (CIE) introduced the two perceptually
based color spaces CIELAB and CIELUV where the latter is typically preferred for 
emissive color technologies such as computer displays. \cite{vcd:Ihaka:2003} discusses
how CIELUV colors can be used for choosing qualitative palettes for statistical graphics such as barplots.
By taking polar coordinates in CIELUV space, it is called HCL (Hue--Chroma--Luminance) space
and qualitative palettes can easily be chosen by using a range of hues for fixed values
of chroma and luminance. Such colors are always balanced towards the same grey 
and thus do not have the problem of varying saturations as the HSV colors.

Here, we discuss how similar ideas can be used for deriving diverging HCL palettes that 
provide a suitable translation of the ideas from the Friendly shading to perceptually
uniform HCL colors. 
The HCL space looks like a double cone with black (zero luminance) at one end and white
at the other (full luminance). In its middle, there is a full color wheel for different
values of chroma (that controls the colorfulness). Unfortunately, the HCL space is not completely
regular as the HSV space is and consequently it cannot be standardized in the same way as the
HSV space. Its dimensions are usually given by a
hue ranging in $[0, 360]$ degrees and chroma and luminance ranging in $[0, 100]$
percent. But not all combinations $(h, c, l)$ yield valid HCL colors and the admissable combinations
of $c$ and $l$ vary across different hues $h$. But for the problem of constructing
a diverging palette, this problem can easily be overcome as we just need two different
hues (a `negative' and a `positive' hue) and hence we can choose two hues that correspond to 
similar shapes in the chroma/luminance plane. The lower panel of Figure~\ref{fig:shadingHSVHCL}
shows two such planes side by side for the hues $h = 0$ and $h = 260$. To obtain a 
sequence of colors with the same properties as the Friendly shading, the palette starts
at a fully saturated red $(0, 100, 50)$, goes via a neutral color, ends at a fully saturated
blue $(260, 100, 50)$, and uses linear interpolation in between. Instead of using white
$(0, 0, 100)$ as the neutral color, a light grey $(0, 0, 90)$ is employed as motivated above.

This diverging palette (see the `significant' colors in Figure~\ref{fig:shadingHSVHCL},
the `non-significant' colors are again discussed in the following section)
uses both chroma, i.e., the colorfulness, and luminance, i.e., 
the amount of grey, to code the absolute size of the quantity visualized, i.e., the
residuals $r_{ij}$ when applied to the independence problem. By changing the neutral
color or by changing the maximum chroma respectively, this can be changed to using only
chroma or luminance for this purpose, but using both seems to be a very effective way
of visualization.

@@

The shading scheme of \cite{vcd:Friendly:1994} was suggested to visualize
the pattern of dependence in contingency tables, as discussed above, but
the presence (or absence) of colors in a plot also always conveys an
impression of interestingness (or un-interstingness respectively). That is,
statisticians might be tempted to interpret the absence of color in a plot
as a clue that there is no significant departure from independence. Or vice
versa, colored cells would convey the impression that there is significant
dependence. Currently, both is not true as can be seen in the left panel
of Figure~\ref{fig:shadings} which shows the mosaic display for the arthritis
data with Friendly shading. Although there is significant dependence, no 
residual exceeds an absolute value of $2$ and hence no cell is colored.
Of course, it can be argued that the shading was not designed for this purpose
and that different cut offs than $2$ and $4$ should be used here. But then
again, it would be nice if these cut offs could be chosen automatically in a
data-driven fashion which is what we do below.

The Friendly shading can be interpreted to be a visualization of the maximum
statistic $M$ from (\ref{eq:MaxAbs}) which always employs the critical values $c_\alpha$
$2$ and $4$. But the problem is that it is not clear to which
significance levels $\alpha$ these critical values correspond because the distribution
of $M$ depends on the underlying contingency table. The natural solution to this
problem is to compute the critical values from the distribution of $M$ in a data-driven
way (i.e., for the table visualized) and use these instead of the hard-coded values
$2$ and $4$. In the right panel of Figure~\ref{fig:shadings} this is done for
the arthritis data by employing the critical values
$\Sexpr{round(as.vector(art_max$qdist(0.9)), digits = 2)}$ at level $\alpha = 0.1$ and
$\Sexpr{round(as.vector(art_max$qdist(0.99)), digits = 2)}$ at level $\alpha = 0.01$
(and using the diverging HCL palette) derived from the permutation distribution of $M$ for 
the arthritis data as described in Section~\ref{sec:tests}. By using these cut offs, the presence of color
in the plot is equivalent to significance (of the maximum statistic $M$) at 
level $\alpha = 0.1$ and $\alpha = 0.01$ respectively and exactly the cells which `cause'
the dependence are highlighted. For the arthritis data, these are in particular the 
cells in the last column that signal that there are significantly more
treated patients and fewer placebo patients with marked improvement than would be expected
under independence.

The significance levels $\alpha = 0.1$ and $0.01$ are chosen because this leads to displays where
fully colored cells are clearly significant (i.e., with a $p$ value below $0.01$),
cells without color are clearly non-significant (i.e., with a $p$ value above $0.1$), and
cells in between can be considered to be weakly significant (i.e., $0.1 \le p < 0.01$).
Of course, users could choose any other set of significance levels they feel comfortable
with, e.g., only a single cutoff at $\alpha = 0.05$ or three cut offs at $0.1$, $0.05$ and $0.01$
etc. Another option could be to use a continuous shading where the $p$ value corresponding
to a cell controls the interpolation between the neutral and the full color. However, our
experience is that this typically results in too much color in a plot which in turn tends
to conceal the important cells and over-emphasizes the unimportant ones. Hence, we stick
to a discrete shading with few colors that is easier to interpret.

This maximum shading is already very flexible and combines visualization and inference.
However, it can only be applied when employing the maximum statistic because it is the only
aggregation functional $\lambda(\cdot)$ where a single large residual $|r_{ij}|$ exceeding
its critical value is equivalent
to a significant value of the whole test statistic $\lambda(\{r_{ij}\})$.\footnote{This
can easily be seen from the following consideration: To obtain a level $\alpha$ test, where 
any residual $|r_{ij}|$ exceeding the critical value $c_\alpha$ gives a significant test
result, the equation $\alpha = P(\mbox{any } |r_{ij}| > c_\alpha)$ must hold. This is equivalent
to $1 - \alpha = P(\mbox{all } |r_{ij}| \le c_\alpha) = P(\max_{i,j} |r_{ij}| \le c_\alpha)$.}
Typically, applying the maximum statistic is feasible and also appropriate for exploratory 
analysis, but it would be desirable to also have a residual-based shading that can incorporate
visualization of significance when the sum-of-squares statistic $X^2$ (or any other functional
$\lambda(\cdot)$) is used. For the reasons discussed above, it is not possible to achieve this by shading
individual cells differently but can only be realized by using different colors for the
whole table. A outlined before, colorfulness is intuitively matched with interestingness,
therefore a rather natural idea is to use the fully colored palette only when the corresponding
test is significant and to use a less colorful palette if not. For the HCL scheme, the
amount of color can conveniently be controlled by varying the maximum chroma value used. For the full
colors, the maximum chroma was set to $100$ as shown in Figure~\ref{fig:shadingHSVHCL} and 
is now reduced to $20$ for the non-significant palette. This palette still codes the absolute
size of the residuals by luminance (i.e., the amount of grey), uses the same neutral grey
for small `un-interesting' residuals, codes positive and negative residuals by different hues,
but gives less emphasis to the pattern by making the plot less colorful. A similar effect
can be obtained in the HSV space if the value is reduced from $1$ to $0.5$ for non-significant 
palettes (see Figure~\ref{fig:shadingHSVHCL}). But as for the significant colors,
the dimensions used for creating this palette are confounded and hence the HCL scheme is clearly
preferable.

\setkeys{Gin}{width=\textwidth}
\begin{figure}[htbp]
\begin{center}
<<shadings,fig=TRUE,echo=FALSE,height=6,width=11,results=hide>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
mosaic(art, newpage = FALSE, gp = shading_Friendly, margins = c(2.5, 4, 2.5, 3), gp_args = list(eps = NULL, lty = 1))
popViewport()
pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
mosaic(art, newpage = FALSE, gp = art_gp, margins = c(2.5, 3, 2.5, 4))
popViewport(2)
@
\caption{Mosaic plot for the Arthritis data with Friendly shading (left) and maximum shading (right).}
\label{fig:shadings}
\end{center}
\end{figure}

\section{Extension to Conditional Independence}

To introduce the new residual-based shadings without too much overhead in
Section~\ref{sec:shadings}, we have only considered the independence problem
in 2-way tables up to now. But, in fact, none of the ideas described are really
confined to the 2-dimensional case as will be outlined in this section.

Mosaic displays have been emphasized in the literature to be an excellent means
of visualization for log-linear models \citep{vcd:Friendly:1999,vcd:Theus+Lauer:1999}.
Problems of interest include complete, joint or conditional independence for all of 
which the tables of expected values and residuals (again Pearson or deviance) can
be computed. \citep{vcd:Friendly:1994,vcd:Friendly:1999} shows that his residual-based 
shading scheme can directly be applied to these more complex independence models.
For inference, the most commonly used aggregation functional for the residuals
is again the sum of squares yielding the associated Pearson or likelihood
ratio statistic respectively \citep{vcd:Agresti:2002}. Hence, our residual-based
shading with different HCL palettes for significant and non-significant results
can directly be applied to these models. For the maximum shading with
data-driven cut offs, even more flexibility is possible than in the 2-way case
when the structure of the independence problem is exploited which will be discussed 
below for conditional independence

Association plots are not commonly used for contingency tables with more than two 
margins, although there is nothing in the definition that would prevent application
in higher dimensions. However, as argued for the mosaic plots by \cite{vcd:Friendly:1999}
and \cite{vcd:Theus+Lauer:1999}, it becomes increasingly
important to choose a good layout as the number of variables grows.
